Predicting trends in the quality of state-of-the-art neural networks without access to training or testing data

### 2020-02-16 submission/background.tex
full pass, fixed lots of typos, etc.; semifinal, nearly done

Statistical mechanics ideas have long had influence on DNN theory and practice~\cite{EB01_BOOK, MM17_TR, BKPx20}; and 
our best-performing metrics (those using fitted PL exponents) are based on statistical mechanics~\cite{MM17_TR, MM18_TR, MM19_HTSR_ICML, MM19_KDD, MM20_SDM}, in particular the recently-developed \emph{Theory of Heavy Tailed Self Regularization (HT-SR)}~\cite{MM18_TR, MM19_HTSR_ICML, MM20_SDM}.  
We emphasize that the way in which we (and HT-SR Theory) \emph{use} statistical mechanics theory is quite different than the way it is more commonly formulated.
Several very good overviews of the more common approach are available~\cite{EB01_BOOK, BKPx20}.
We \emph{use} statistical mechanics in a broader sense, drawing upon techniques from quantitative finance and random matrix theory.  % (RMT).
Thus, much more relevant for our metholodogical approach is older work of Bouchaud, Potters, Sornette, and coworkers~\cite{BouchaudPotters03, SornetteBook, BP11, bun2017} on the statistical mechanics of heavy tailed and strongly correlated systems.
% Probably best not to cite at this point %\charles{, and current work (in progress)\cite{blog}}

@@

%% \paragraph{Practical problems need a practical theory.}
%% \nred{Exiting ML theory does not address many practical AI problems that our techiques directly target.}
%% %There are many very practical problems in ML that are poorly addressed by existing theory and that either motivated our work or should be addressable by our techniques.
%% Here are several.
%% \begin{itemize}[leftmargin=*]
%% \item
%% \textbf{Simplicity, or lack of,  accuracy metrics.}
%% \michael{XXX.  Put in two or at most three sentences.}
%% \item
%% \textbf{Information leakage in the production pipeline.}
%% \michael{XXX.  Put in two or at most three sentences.}
%% \item
%% \textbf{Cost of acquring labeled data.}
%% \michael{XXX.  Put in two or at most three sentences.}
%% \end{itemize}
%% 
%% \charles{see comments in tex file}\michael{Let's touch base, not sure what to do.}
%% % we adress this above
%% %Importantly, there are many examples in ML where (as a practical matter) there is no reliable notion of accuracy, e.g., when generating fake text or realiztic chatbots, developing self driving cars, or just clusterting user profiles.
%% %\footnote{For example, current chatbots use perplexity as a proxy for passing a Turing test.}
%% %, and when distilling a reliable model in some way to obtain comparable training/test quality but that damages the model in some other subtle way. 
%% %\michael{That last example is awkward.  It would be good to have a better example and plant seeds for model distillation elsewhere.}

>>>>

Statistical mechanics ideas have long had influence on DNN theory and practice~\cite{EB01_BOOK, MM17_TR, BKPx20}; and 
our best-performing metrics (those using fitted PL exponents) are based on statistical mechanics~\cite{MM17_TR, MM18_TR, MM19_HTSR_ICML, MM19_KDD, MM20_SDM}, in particular the recently-developed \emph{Theory of Heavy Tailed Self Regularization (HT-SR)}~\cite{MM18_TR, MM19_HTSR_ICML, MM20_SDM}.  
We emphasize that the way in which we (and HT-SR Theory) \emph{use} statistical mechanics theory is quite different than the way it is more commonly formulated.
Several very good overviews of the more common approach are available~\cite{EB01_BOOK, BKPx20}.
We \emph{use} statistical mechanics in a broader sense, drawing upon techniques from quantitative finance and random matrix theory.  % (RMT).
Thus, much more relevant for our methodological approach is older work of Bouchaud, Potters, Sornette, and coworkers~\cite{BouchaudPotters03, SornetteBook, BP11, bun2017} on the statistical mechanics of heavy tailed and strongly correlated systems.
% Probably best not to cite at this point %\charles{, and current work (in progress)\cite{blog}}

@@

%% \paragraph{Practical problems need a practical theory.}
%% \nred{Exiting ML theory does not address many practical AI problems that our techniques directly target.}
%% %There are many very practical problems in ML that are poorly addressed by existing theory and that either motivated our work or should be addressable by our techniques.
%% Here are several.
%% \begin{itemize}[leftmargin=*]
%% \item
%% \textbf{Simplicity, or lack of,  accuracy metrics.}
%% \michael{XXX.  Put in two or at most three sentences.}
%% \item
%% \textbf{Information leakage in the production pipeline.}
%% \michael{XXX.  Put in two or at most three sentences.}
%% \item
%% \textbf{Cost of acquiring labeled data.}
%% \michael{XXX.  Put in two or at most three sentences.}
%% \end{itemize}
%% 
%% \charles{see comments in tex file}\michael{Let's touch base, not sure what to do.}
%% % we address this above
%% %Importantly, there are many examples in ML where (as a practical matter) there is no reliable notion of accuracy, e.g., when generating fake text or realistic chatbots, developing self driving cars, or just clustering user profiles.
%% %\footnote{For example, current chatbots use perplexity as a proxy for passing a Turing test.}
%% %, and when distilling a reliable model in some way to obtain comparable training/test quality but that damages the model in some other subtle way. 
%% %\michael{That last example is awkward.  It would be good to have a better example and plant seeds for model distillation elsewhere.}

### 2020-02-16 submission/conclusion.tex
full pass, fixed lots of typos, etc.; semifinal, nearly done

We have developed (based on strong theory) and evaluated (on a large corpus of publicly-available pretrained models from CV and NLP) methods to predict trends in the quality of state-of-the-art neural networks---without access to training or testing data.
Prior to our work, it was not obvious that norm-based metrics would perform well to predict trends in quality \emph{across} models (as they are usually used \emph{within} a given model or parameterized model class, e.g., to bound generalization error or to construct regularizers).
Our results are the first to demonstrate that they can be used for this important practical problem.
That PL-based metrics perform better (than norm-based metrics) should not be surprising---at least to those familiar with the statisical mechanics of heavy tailed and strongly correlated systems~\cite{BouchaudPotters03, SornetteBook, BP11, bun2017} (since our use of PL exponents is designed to capture the idea that well-trained models capture correlations over many size scales in the data).
Again, though, our results are the first to demonstrate this.
It is also gratifying that our approach can be used to provide fine-scale insight (such as rationalizing the flow of correlations or the collapse of size scale) throughout a network.

We conclude with a few comments on what a \emph{practical theory} of DNNs should look like.
To do so, we distinguish between two types of theories:
\emph{non-empirical or analogical theories}, in which one creates, often from general principles, a very simple toy model that can be analyzed rigorously, and one then argues that the model is relevant to the system of interest; and 
\emph{semi-empirical theories}, in which there exists a rigorous asymptotic theory, which comes with parameters, for the system of interest, and one then adjusts or fits those parameters to the finite non-asymptotic data.
A drawback of the former approach is that it typically makes very strong assumptions on the data, and the strength of those assumptions can limit the practical applicability of the theory.
Nearly all of the work on the theory of DNNs focuses on the former type of theory.
Our approach focuses on the latter type of theory.
Our results, which are based on our \emph{use} of sophisticated statistical mechanics theory to solve important practical DNN problems, suggests that our approach should be of interest more generally for those interested in developing a practical DNN theory.

>>>>

We have developed (based on strong theory) and evaluated (on a large corpus of publicly-available pretrained models from CV and NLP) methods to predict trends in the quality of state-of-the-art neural networks---without access to training or testing data.
Prior to our work, it was not obvious that norm-based metrics would perform well to predict trends in quality \emph{across} models (as they are usually used \emph{within} a given model or parameterized model class, e.g., to bound generalization error or to construct regularizers).
Our results are the first to demonstrate that they can be used for this important practical problem.
That PL-based metrics perform better (than norm-based metrics) should not be surprising---at least to those familiar with the statistical mechanics of heavy tailed and strongly correlated systems~\cite{BouchaudPotters03, SornetteBook, BP11, bun2017} (since our use of PL exponents is designed to capture the idea that well-trained models capture correlations over many size scales in the data).
Again, though, our results are the first to demonstrate this.
It is also gratifying that our approach can be used to provide fine-scale insight (such as rationalizing the flow of correlations or the collapse of size scale) throughout a network.

We conclude with a few comments on what a \emph{practical theory} of DNNs should look like.
To do so, we distinguish between two types of theories:
\emph{non-empirical or analogical theories}, in which one creates, often from general principles, a very simple toy model that can be analyzed rigorously, and one then argues that the model is relevant to the system of interest; and 
\emph{semi-empirical theories}, in which there exists a rigorous asymptotic theory, which comes with parameters, for the system of interest, and one then adjusts or fits those parameters to the finite non-asymptotic data.
A drawback of the former approach is that it typically makes very strong assumptions on the data, and the strength of those assumptions can limit the practical applicability of the theory.
Nearly all of the work on the theory of DNNs focuses on the former type of theory.
Our approach focuses on the latter type of theory.
Our results, which are based on \emph{using} sophisticated statistical mechanics theory and \emph{solving} important practical DNN problems, suggests that the latter approach should be of interest more generally for those interested in developing a \emph{practical DNN~theory}.

### 2020-02-16 submission/cv_models.tex
full pass, fixed lots of typos, etc.; semifinal, nearly done

We have examined the performance of the four quality metrics (Log Frobenius norm, Log Spectral norm, Weighted Alpha, and Log $\alpha$-Norm) applied to each of the VGG, ResNet, ResNet-1K, and DenseNet series.
To start, Figure~\ref{fig:vgg-metrics} considers the VGG series (in particular, the pretrained models VGG11, VGG13, VGG16, and VGG19, with and without BN), and it plots the four quality metrics versus the reported test accuracies~\cite{pytorch},% 
\footnote{That is, these test accuracies have been previously reported and made publicly-avaiable by others.  We take them as given, and we do not attempt to reproduce/verify them, since we do not permit ourselves any access to training/test data.}
as well as a basic linear regression line. 
%with the Root Mean Squared Error (RMSE) shown, \michael{XXX.  WHERE IS RMSE.  IN THAT TABLE?}
All four metrics correlate quite well with the reported Top1 accuracies, with smaller norms and smaller values of $\hat{\alpha}$ implying better generalization (i.e., greater accuracy, lower error). 
While all four metrics perform well, notice that the Log $\alpha$-Norm metric ($\log\Vert\mathbf{W}\Vert_{\alpha}^{\alpha}$) performs best (with an RMSE of $0.42$, see Table~\ref{table:cv-models}); and the Weighted Alpha metric ($\hat\alpha =\alpha\log\lambda_{max} $), which is an approximation to the Log $\alpha$-Norm metric~\cite{MM20_unpub_work}, performs second best (with an RMSE of $0.48$, see Table~\ref{table:cv-models}).

@@

    \subfigure[ ResNet, Log $\alpha$-Norm ]{
        \includegraphics[width=4.9cm]{img/ResNet_logpnorm_accs.png}
        %KDD% \includegraphics[width=4.0cm]{img/ResNet_logpnorm_accs.png}
        \label{fig:resnet-accuracy}
    }
    %KDD% 
    \qquad
    \subfigure[ ResNet-1K, Log $\alpha$-Norm ]{
        \includegraphics[width=4.9cm]{img/ResNet-1K_logpnorm_accs.png}
        %KDD% \includegraphics[width=4.0cm]{img/ResNet-1K_logpnorm_accs.png}
        \label{fig:resnet1k-accuracy}
    }
    %\qquad
    %\subfigure[ DenseNet, Log $\alpha$-Norm ]{
    %    %\includegraphics[width=4.4cm]{img/DenseNet_logpnorm_accs.png}
    %    \includegraphics[width=4.0cm]{img/DenseNet_logpnorm_accs.png}
    %    \label{fig:densenet-accuracy}
    %}
    \caption{Comparison of Avergage $\alpha$-Norm quality metric ($\langle\log\Vert\mathbf{X}\Vert_{\alpha}^{\alpha}\rangle$) versus reported Top1 test accuracy for the ResNet and ResNet-1K pretrained (pyTorch) models. }
%, and DenseNet models. }The corresponding results for VGG are shown in Figure~\ref{fig:vgg-pnorm}.  }
    \label{fig:cv2-accuracy}
\end{figure}

%MM% \begin{figure}
%MM% \centering
%MM% \begin{minipage}[b]{.22\textwidth}
%MM% %<Code for the first figure>
%MM% XXX
%MM% \caption{Caption}\label{label-a}
%MM% \end{minipage}\qquad
%MM% \begin{minipage}[b]{.22\textwidth}
%MM% %<Code for the second figure>
%MM% XXX
%MM% \caption{Caption}\label{label-b}
%MM% \end{minipage}
%MM% \end{figure}

@@

The DenseNet series has similar behavior to what we see in Figures~\ref{fig:vgg-metrics} and~\ref{fig:cv2-accuracy} for the other models.
However, as noted in Table~\ref{table:cv-models}, it has only 4 data points.  
In our larger analysis, in Section~\ref{sxn:all_cv_models}, we will only include series with 5 or more models.
%For completemnss and reproducibility here, the DenseNet Log $\alpha$-Norm plot is given in Appendix~\ref{sxn:appendix}.
(Note that these and many other such plots can be seen on our publicly-available~repo.)

@@

%KDD% \vspace{-1mm}
\paragraph{Comparison of VGG, ResNet, and DenseNet Architectures.}
We can interpret these observations by recalling the architectural differences between the VGG, ResNet, and DenseNet architectures,
and, in particular, the number of of residual connections. VGG resembles the  traditional convolutional architectures,
such as LeNet5, and consists of several [Conv2D-Maxpool-ReLu] blocks, followed by 3 large Fully Connected (FC) layers.
ResNet greatly improved on VGG by replacing the large FC layers, shrinking the Conv2D blocks, and introducing \emph{residual connections}.
This optimized approach allows for greater accurary with far fewer parameters (and GPU memory requirements), and
ResNet models of up to 1000 layers have been trained~\cite{resnet1000}.

We conjuecture that the effeciency and effectivness of ResNet is reflected in the smaller and more stable $\alpha\sim 2.0$, across nearly all layers, indicating that the inner layers are very well correlated and strongly optimized.
Contrast this with the DenseNet models, which contains many connections between every layer.
Our results (large $\alpha$, meaning they even a PL model is probably a poor fit) suggest that DenseNet has too many connections, diluting high quality interactions across layers, and leaving many layers very poorly optimized.

@@

More generally, we can understand the results presented in Figure~\ref{fig:3models-alpha-layers} in terms of what we will call the \emph{Correlation Flow} of the model. 
Recall that the average Log $\alpha$-Norm metric and the Weighted Alpha metric are based on HT-SR Theory~\cite{MM18_TR, MM19_HTSR_ICML, MM20_SDM}, which is in turn based on ideas from the statisical mechanics of heavy tailed and strongly correlated systems~\cite{BouchaudPotters03, SornetteBook, BP11, bun2017}. 
There, one expects the weight matrices of well-trained DNNs will exhibit correlations over many size scales. 
Their ESDs can be well-fit by a (truncated) PL, with exponents $\alpha\in[2,4]$.
Much larger values $(\alpha\gg 6)$ may reflect poorer PL fits, whereas smaller values $(\alpha\sim 2)$, are associated with models that generalize better.
%, but the presence of many adjacent layers with relatively good PL fits \emph{suggests} that those adjacent layers ``impedance match'' well and that the correlations that are captured by one layer propagate well to subsequent layers.
Informally, one would expect a DNN model to perform well when it fascilitates the propagation of information/features across layers.
Previous work argues this by computing the gradients over the input data.
In the absence of training/test data, one might hope that this leaves empirical signatures on weight matrices, and thus we can to try to quantify this by measuring the PL properties of weight matrices.
In this case, smaller $\alpha$ values correspond to layers in which correlations across multiple scales are better captured~\cite{MM18_TR,SornetteBook}, and we expect that small $\alpha$ values that are stable across multiple layers enable better \emph{correlation flow} through the network.
We have seen this in many models, including those shown in Figure~\ref{fig:3models-alpha-layers}.

@@

The similarity between norm-based metrics and PL-based metrics suggests a question: is the Weighted Alpha metric just a variation of the more familiar norm-based metrics?  %%, or does it capture something different?  
More generally, do fitted $\alpha$ values contain information not captured by norms? 
In examining hundreds of pretrained models, we have found several anamolies that demonsrate the power of our approach.
In particular, to show that $\alpha$ does capture something different, consider the following example, which looks at a compressed/distilled DNN model~\cite{CWZZ17_TR}.
In this example, we show that some distillation methods may actually \emph{break} models unexpectedly by introducing what we call \emph{Scale Collapse}, where several distilled layers have unexpectedly small Spectral Norms.

We consider ResNet20, trained on CIFAR10, before and after applying the Group Regularization distillation technique, as implemented in the \texttt{distiller} package~\cite{distiller}.
We analyze the pretrained 4D\_regularized\_5Lremoved baseline and fine-tuned models. 
The reported baseline test accuracies (Top1$=91.45$ and Top5$=99.75$) are better than the reported fine-tuned test accuracies (Top1$=91.02$ and Top5$=99.67$).  Because the baseline accuracy is greater,  the previous results on ResNet (Table~\ref{table:cv-models} and Figure~\ref{fig:cv2-accuracy}) suggest that the baseline Spectral Norms should be \emph{smaller} on average than the fine-tuned ones. \emph{The oppopsite is observed.}
%\michael{Do we have something backwards here: where we observed that the average norm to increase with decreasing test error (as it ``should'' not), whereas the average PL exponent $\alpha$ decreases (as ``expected'' from HT-SR Theory).  }
%In both cases (Frobenius norm results not shown), we observe the opposite.
%%\charles{WHy is this confusing? Call me...}
Figure~\ref{fig:resnet204D5L} presents the Spectral Norm (here denoted $\log\lambda_{max}$) and PL exponent ($\alpha$) for each individual layer weight matrix $\mathbf{W}$.%
\footnote{Here, we only include layer matrices or feature maps with $M\ge50$.}
On the other hand, the $\alpha$ values (in Figure~\ref{fig:resnet204Dalpha}) do not differ systematically between the baseline and fine-tuned models.
%\michael{Some comment about big jump in norm; but fix previous backwards question first.}--SEE BELOW
Also (not shown), the average (unweighted) baseline $\alpha$ is \emph{smaller} than the fine-tuned average (as predicted by HT-SR Theory, the basis of $\hat{\alpha}$).

That being said, Figure~\ref{fig:resnet204Dalpha} also depcits two very large $\alpha\gg 6$ values for the baseline, but not for the fine-tuned, model.
This suggests the baseline model has at least two over-parameterized/under-trained layers, and that the distillation method does, in fact, improve the fine-tuned model by compressing these layers.

>>>>

We have examined the performance of the four quality metrics (Log Frobenius norm, Log Spectral norm, Weighted Alpha, and Log $\alpha$-Norm) applied to each of the VGG, ResNet, ResNet-1K, and DenseNet series.
To start, Figure~\ref{fig:vgg-metrics} considers the VGG series (in particular, the pretrained models VGG11, VGG13, VGG16, and VGG19, with and without BN), and it plots the four quality metrics versus the reported test accuracies~\cite{pytorch},% 
\footnote{That is, these test accuracies have been previously reported and made publicly-available by others.  We take them as given, and we do not attempt to reproduce/verify them, since we do not permit ourselves any access to training/test data.}
as well as a basic linear regression line. 
All four metrics correlate quite well with the reported Top1 accuracies, with smaller norms and smaller values of $\hat{\alpha}$ implying better generalization (i.e., greater accuracy, lower error). 
While all four metrics perform well, notice that the Log $\alpha$-Norm metric ($\log\Vert\mathbf{W}\Vert_{\alpha}^{\alpha}$) performs best (with an RMSE of $0.42$, see Table~\ref{table:cv-models}); and the Weighted Alpha metric ($\hat\alpha =\alpha\log\lambda_{max} $), which is an approximation to the Log $\alpha$-Norm metric~\cite{MM20_unpub_work}, performs second best (with an RMSE of $0.48$, see Table~\ref{table:cv-models}).

@@

    \subfigure[ ResNet, Log $\alpha$-Norm ]{
        \includegraphics[width=4.9cm]{img/ResNet_logpnorm_accs.png}
        %KDD% \includegraphics[width=4.0cm]{img/ResNet_logpnorm_accs.png}
        \label{fig:resnet-accuracy}
    }
    %KDD% 
    \qquad
    \subfigure[ ResNet-1K, Log $\alpha$-Norm ]{
        \includegraphics[width=4.9cm]{img/ResNet-1K_logpnorm_accs.png}
        %KDD% \includegraphics[width=4.0cm]{img/ResNet-1K_logpnorm_accs.png}
        \label{fig:resnet1k-accuracy}
    }
    %\qquad
    %\subfigure[ DenseNet, Log $\alpha$-Norm ]{
    %    %\includegraphics[width=4.4cm]{img/DenseNet_logpnorm_accs.png}
    %    \includegraphics[width=4.0cm]{img/DenseNet_logpnorm_accs.png}
    %    \label{fig:densenet-accuracy}
    %}
    \caption{Comparison of Average $\alpha$-Norm quality metric ($\langle\log\Vert\mathbf{X}\Vert_{\alpha}^{\alpha}\rangle$) versus reported Top1 test accuracy for the ResNet and ResNet-1K pretrained (pyTorch) models. }
%, and DenseNet models. }The corresponding results for VGG are shown in Figure~\ref{fig:vgg-pnorm}.  }
    \label{fig:cv2-accuracy}
\end{figure}

@@

The DenseNet series has similar behavior to what we see in Figures~\ref{fig:vgg-metrics} and~\ref{fig:cv2-accuracy} for the other models.
However, as noted in Table~\ref{table:cv-models}, it has only 4 data points.  
In our larger analysis, in Section~\ref{sxn:all_cv_models}, we will only include series with 5 or more models.
%For completeness and reproducibility here, the DenseNet Log $\alpha$-Norm plot is given in Appendix~\ref{sxn:appendix}.
(Note that these and many other such plots can be seen on our publicly-available~repo.)

@@

%KDD% 
\vspace{-1mm}
\paragraph{Comparison of VGG, ResNet, and DenseNet Architectures.}
We can interpret these observations by recalling the architectural differences between the VGG, ResNet, and DenseNet architectures,
and, in particular, the number of of residual connections. VGG resembles the  traditional convolutional architectures,
such as LeNet5, and consists of several [Conv2D-Maxpool-ReLu] blocks, followed by 3 large Fully Connected (FC) layers.
ResNet greatly improved on VGG by replacing the large FC layers, shrinking the Conv2D blocks, and introducing \emph{residual connections}.
This optimized approach allows for greater accuracy with far fewer parameters (and GPU memory requirements), and
ResNet models of up to 1000 layers have been trained~\cite{resnet1000}.

We conjecture that the efficiency and effectiveness of ResNet is reflected in the smaller and more stable $\alpha\sim 2.0$, across nearly all layers, indicating that the inner layers are very well correlated and strongly optimized.
Contrast this with the DenseNet models, which contains many connections between every layer.
Our results (large $\alpha$, meaning they even a PL model is probably a poor fit) suggest that DenseNet has too many connections, diluting high quality interactions across layers, and leaving many layers very poorly optimized.

@@

More generally, we can understand the results presented in Figure~\ref{fig:3models-alpha-layers} in terms of what we will call the \emph{Correlation Flow} of the model. 
Recall that the average Log $\alpha$-Norm metric and the Weighted Alpha metric are based on HT-SR Theory~\cite{MM18_TR, MM19_HTSR_ICML, MM20_SDM}, which is in turn based on ideas from the statistical mechanics of heavy tailed and strongly correlated systems~\cite{BouchaudPotters03, SornetteBook, BP11, bun2017}. 
There, one expects the weight matrices of well-trained DNNs will exhibit correlations over many size scales. 
Their ESDs can be well-fit by a (truncated) PL, with exponents $\alpha\in[2,4]$.
Much larger values $(\alpha\gg 6)$ may reflect poorer PL fits, whereas smaller values $(\alpha\sim 2)$, are associated with models that generalize better.
%, but the presence of many adjacent layers with relatively good PL fits \emph{suggests} that those adjacent layers ``impedance match'' well and that the correlations that are captured by one layer propagate well to subsequent layers.
Informally, one would expect a DNN model to perform well when it facilitates the propagation of information/features across layers.
Previous work argues this by computing the gradients over the input data.
In the absence of training/test data, one might hope that this leaves empirical signatures on weight matrices, and thus we can to try to quantify this by measuring the PL properties of weight matrices.
In this case, smaller $\alpha$ values correspond to layers in which correlations across multiple scales are better captured~\cite{MM18_TR,SornetteBook}, and we expect that small $\alpha$ values that are stable across multiple layers enable better \emph{correlation flow} through the network.
We have seen this in many models, including those shown in Figure~\ref{fig:3models-alpha-layers}.

@@

The similarity between norm-based metrics and PL-based metrics suggests a question: is the Weighted Alpha metric just a variation of the more familiar norm-based metrics?  %%, or does it capture something different?  
More generally, do fitted $\alpha$ values contain information not captured by norms? 
In examining hundreds of pretrained models, we have found several anomalies that demonstrate the power of our approach.
In particular, to show that $\alpha$ does capture something different, consider the following example, which looks at a compressed/distilled DNN model~\cite{CWZZ17_TR}.
In this example, we show that some distillation methods may actually \emph{break} models unexpectedly by introducing what we call \emph{Scale Collapse}, where several distilled layers have unexpectedly small Spectral Norms.

We consider ResNet20, trained on CIFAR10, before and after applying the Group Regularization distillation technique, as implemented in the \texttt{distiller} package~\cite{distiller}.
We analyze the pretrained 4D\_regularized\_5Lremoved baseline and fine-tuned models. 
The reported baseline test accuracies (Top1$=91.45$ and Top5$=99.75$) are better than the reported fine-tuned test accuracies (Top1$=91.02$ and Top5$=99.67$).  Because the baseline accuracy is greater,  the previous results on ResNet (Table~\ref{table:cv-models} and Figure~\ref{fig:cv2-accuracy}) suggest that the baseline Spectral Norms should be \emph{smaller} on average than the fine-tuned ones. \emph{The opposite is observed.}
%\michael{Do we have something backwards here: where we observed that the average norm to increase with decreasing test error (as it ``should'' not), whereas the average PL exponent $\alpha$ decreases (as ``expected'' from HT-SR Theory).  }
%In both cases (Frobenius norm results not shown), we observe the opposite.
%%\charles{WHy is this confusing? Call me...}
Figure~\ref{fig:resnet204D5L} presents the Spectral Norm (here denoted $\log\lambda_{max}$) and PL exponent ($\alpha$) for each individual layer weight matrix $\mathbf{W}$.%
\footnote{Here, we only include layer matrices or feature maps with $M\ge50$.}
On the other hand, the $\alpha$ values (in Figure~\ref{fig:resnet204Dalpha}) do not differ systematically between the baseline and fine-tuned models.
%\michael{Some comment about big jump in norm; but fix previous backwards question first.}--SEE BELOW
Also (not shown), the average (unweighted) baseline $\alpha$ is \emph{smaller} than the fine-tuned average (as predicted by HT-SR Theory, the basis of $\hat{\alpha}$).

That being said, Figure~\ref{fig:resnet204Dalpha} also depicts two very large $\alpha\gg 6$ values for the baseline, but not for the fine-tuned, model.
This suggests the baseline model has at least two over-parameterized/under-trained layers, and that the distillation method does, in fact, improve the fine-tuned model by compressing these layers.

### 2020-02-16 submission/introduction.tex
full pass, fixed lots of typos, etc.; semifinal, nearly done

A common problem in machine learning (ML) 
%and artificial intelligence (AI) 
is to evaluate the quality of a given model.
A popular way to accomplish this
%, in particular in academic environments, 
is to train a model and then evaluate its training/testing error.
There are many problems with this approach.
The training/testing curves give very limited insight into the overall properties of the model; 
they do not take into account the (often large human and CPU/GPU) time for hyperparameter fiddling;
they typically do not correlate with other properties of interest such as robustness or fairness or interpretability; 
and so on.
A less well-known problem, but one that is increasingly important, in particular in industrial-scale artificial intelligence (AI), arises when the model \emph{user} is not the model \emph{developer}.
Here, one may not have access to either the training data or the testing data.
Instead, one may simply be given a model that has already been trained---a \emph{pretrained model}---and need to use it ``as is,'' or to fine-tune and/or compress it and then use it.

@@

Most importantly, in industry, one faces unique practical problems such as: do we have enough data for this model? 
Indeed, high quality, labeled data can be very expensive to acquire, and this cost can make or break a project.
Methods that are developed and evaluated on any well-defined publicly-available coprus of data, no matter how large or diverse or interesting, are clearly not going to be well-suited to address problems such as this.
It is of great practical interest to have metrics to evaluate the quality of a trained model---in the absence of training/testing data and without any detailed knowledge of the training/testing process.  
We seek a practical theory for pretrained models which can predict how, when, and why such models can be expected to perform well or~poorly.

@@

The \emph{quality metrics} we consider are based on the spectral properties of the layer weight matrics.
They are based on norms of weight matrices (such norms have been used in traditional statistical learning theory to bound capacity and construct regularizers) and/or parameters of power law (PL) fits of the eigenvalues of weight matrices (such PL fits are based on statistical mechanics approaches to DNNs).
Note that, while we use traditional norm-based and PL-based metrics, our goals are not the traditional goals.
Unlike more common ML approaches, \emph{we do not seek a bound on the generalization} (e.g., by evaluating training/test error during training), \emph{we do not seek a new regularizer}, and \emph{we do not aim to evaluate a single model} (e.g., as with hyperparmeter optimization).% 
\footnote{One could of course use these techniques to improve training, and we have been asked about that, but we are not interested in that here. Our main goal here is to use these techniques to evaluate properties of state-of-the-art pretrained DNN models.}
Instead, we want to examine different models across common architecture series, and we want to compare models between different architectures themselves, and in both cases, we ask:
\begin{quote}
\emph{Can we predict trends in the quality of pretrained DNN models without access to training or testing data?}  
\end{quote}

@@

We start in Section~\ref{sxn:background} and Section~\ref{sxn:methods} with background and an overview of our general approach.
In Section \ref{sxn:cv}, we study three well-known widely-available DNN CV architectures (the VGG, ResNet, and DenseNet series of models); and we provide an illustration of our basic methodology, both to evaluate the different metrics against reported test accuracies and to use quality metrics to understand model properties.
Then, in Section~\ref{sxn:nlp}, we look at several variations of a popular NLP DNN architecture (the OpenAI GPT and GPT2 models); and we show how model quality and properties vary between several variants of GPT and GPT2, including how metrics behave similarly and differently.
Then, in Section \ref{sxn:all_cv_models}, we present results based on an analysis of hundreds of pretraind DNN models, showing how well each metric predicts the reported test accuracies, and how the PL-based metrics perform remarkably well.
Finally, in Section~\ref{sxn:conc}, we provide a brief discussion and conclusion.

>>>>

A common problem in machine learning (ML) 
%and artificial intelligence (AI) 
is to evaluate the quality of a given model.
A popular way to accomplish this
%, in particular in academic environments, 
is to train a model and then evaluate its training/testing error.
There are many problems with this approach.
The training/testing curves give very limited insight into the overall properties of the model; 
they do not take into account the (often large human and CPU/GPU) time for hyperparameter fiddling;
they typically do not correlate with other properties of interest such as robustness or fairness or interpretability; 
and so on.
A less well-known problem, but one that is increasingly important, in particular in industrial-scale artificial intelligence (AI), arises when the model \emph{user} is not the model \emph{developer}.
Here, one may not have access to either the training data or the testing data.
Instead, one may simply be given a model that has already been trained---a \emph{pretrained model}---and need to use it as-is, or to fine-tune and/or compress it and then use it.

@@

Most importantly, in industry, one faces unique practical problems such as: do we have enough data for this model? 
Indeed, high quality, labeled data can be very expensive to acquire, and this cost can make or break a project.
Methods that are developed and evaluated on any well-defined publicly-available corpus of data, no matter how large or diverse or interesting, are clearly not going to be well-suited to address problems such as this.
It is of great practical interest to have metrics to evaluate the quality of a trained model---in the absence of training/testing data and without any detailed knowledge of the training/testing process.  
We seek a practical theory for pretrained models which can predict how, when, and why such models can be expected to perform well or~poorly.

@@

The \emph{quality metrics} we consider are based on the spectral properties of the layer weight matrices.
They are based on norms of weight matrices (such norms have been used in traditional statistical learning theory to bound capacity and construct regularizers) and/or parameters of power law (PL) fits of the eigenvalues of weight matrices (such PL fits are based on statistical mechanics approaches to DNNs).
Note that, while we use traditional norm-based and PL-based metrics, our goals are not the traditional goals.
Unlike more common ML approaches, \emph{we do not seek a bound on the generalization} (e.g., by evaluating training/test error during training), \emph{we do not seek a new regularizer}, and \emph{we do not aim to evaluate a single model} (e.g., as with hyperparameter optimization).% 
\footnote{One could of course use these techniques to improve training, and we have been asked about that, but we are not interested in that here. Our main goal here is to use these techniques to evaluate properties of state-of-the-art pretrained DNN models.}
Instead, we want to examine different models across common architecture series, and we want to compare models between different architectures themselves, and in both cases, we ask:
\begin{quote}
\emph{Can we predict trends in the quality of pretrained DNN models without access to training or testing data?}  
\end{quote}

@@

We start in Section~\ref{sxn:background} and Section~\ref{sxn:methods} with background and an overview of our general approach.
In Section \ref{sxn:cv}, we study three well-known widely-available DNN CV architectures (the VGG, ResNet, and DenseNet series of models); and we provide an illustration of our basic methodology, both to evaluate the different metrics against reported test accuracies and to use quality metrics to understand model properties.
Then, in Section~\ref{sxn:nlp}, we look at several variations of a popular NLP DNN architecture (the OpenAI GPT and GPT2 models); and we show how model quality and properties vary between several variants of GPT and GPT2, including how metrics behave similarly and differently.
Then, in Section \ref{sxn:all_cv_models}, we present results based on an analysis of hundreds of pretrained DNN models, showing how well each metric predicts the reported test accuracies, and how the PL-based metrics perform remarkably well.
Finally, in Section~\ref{sxn:conc}, we provide a brief discussion and conclusion.
